---
title: "Functions and arguments"
output: github_document
---

[<<< Previous](01-introduction.md) | [Next >>>](03-objects.md)

Note from Connor:
I think the function that we have them construct should just be their version of the ```mean()``` function.
```{r}
take_avg <- function(x){
  
  # this function takes the same of all values in x. Assumes x is numerical
  s <- sum(x)
  
  # this counts the number of elements in x
  l <- length(x)
  
  # we're dividing the sum of values by the number of values to calculate the mean
  avg <- s / l
  
  # this specifies the value to be returned by the function
  return(avg)
}
```


# Functions
When we want to make something happen in R, we can call a function. Most functions will produce results that you can see immediately, such as the square root function in R. The function ```sqrt()``` is part of the base R package that comes with your installation of R.
```{r}
sqrt(16)
```

## Code body
Typing the name of the function sans the paranthesis (as it is seen above), e.g. just ```mean```, will return the code that is stored in the function.
```{r}
mean
```

## Run a function
To run the function, you will have to add the paranthesis behind the function name, e.g. ```sum()```. If an input is needed for the function to run, you will insert in between the paranthesis. Different functions can require different kinds of input.
```{r}
sum(1,3,7,9,4)
```

## Help page
Adding a ? in front of a function will open up the documentation for that particular function. It can help explain what that function can do and serves as a pretty useful help page when you are learning or troubleshooting.
```{r}
?sum
```

# Arguments
The input you provide for your function are called arguments. Some functions take a single argument while others take multiple.
```{r}
#Example of a function that takes a single argument:
factorial(3)

#Example of a function that takes multiple arguments:
paste("Hello", " World!", sep = "")
```

## args()
This ```args()``` function tell you what arguments a function recognizes. 
```{r}
#For example:
args(sum)
```
From the output of ```args(sum)```, we know that this function takes 2 arguments. The names of these inputs are helpful when we are looking them up in the documentation page. For example, in the case of ```?sum``` we can find out from the documentation that the argument ```na.rm``` is asking if missing values should be removed when this funciton is called. 

## Matching arguments
You can also use the names of the arguments to specify the input within the function. If you don't use argument names, R will match your inputs based on the order that they are written. 
```{r}
log(100, 10) 
#is different from 
log(10, 100)
#specifying your input with the argument names can help avoid potential errors
log(x = 100, base = 10)
log(base = 100, x = 10)
```

## Default arguments
Some arguments are optional, such as ```base``` in ```log()```. In the case of an optional argument, you need not supply a value and the function will still run. In this case, the ```base``` argument has a default value of exp(1) and if you do not provide an alternative base value it will use its default value. However, the argument of x is not optional and requires an input or it will not run.




Tutorial based off [R Studio Primers](https://rstudio.cloud/learn/primers)

[<<< Previous](01-introduction.md) | [Next >>>](03-objects.md)   
[Glossary](glossary.md)