---
title: "Data types"
output: github_document
---

[\<\<\< Previous](04-vectors.md) | [Next \>\>\>](06-data-structure.md)

## Data types

R interprets the type of your vector based on the characteristics of its values. For instance, R can do math with vectors that contain numerical data:

```{r}
nums <- c(1, 2, 3)

nums * 2
```

If you have a vector of character strings, R can count the number of characters in each string:

```{r}
sentence <- c("I", "love", "to", "code", "in", "R")

nchar(sentence)
```

Each *data type* has its own unique properties.

Here is a list of the different data types:

* ``int`` stands for integers, or whole numbers.

* ``dbl`` stands for doubles, or numbers with decimals.

* ``chr`` stands for character vectors, or strings.

* ``dttm`` stands for date-times (a date + a time).

* ``lgl`` stands for logical, vectors that contain only TRUE or FALSE.

* ``fctr`` stands for factors, which R uses to represent categorical variables with fixed possible values.

* ``date`` stands for dates.

## typeof()



## Getting it right

Vectors cannot contain more than one type of data. If you try, R will coerce your vector to a single data type that you weren't intending.

```{r}
mix_vec <- c(1, "two", 3L)

mix_vec

typeof(mix_vec)
```


[\<\<\< Previous](04-vectors.md) | [Next \>\>\>](06-data-structure.md)  
[Glossary](glossary.md)
