---
title: "Data types"
output: github_document
---

[\<\<\< Previous](04-vectors.md) | [Next \>\>\>](06-data-structure.md)

## Data types

R interprets the type of your vector based on the characteristics of its values. For instance, R can do math with vectors that contain numerical data:

```{r}
nums <- c(1, 2, 3)

nums * 2
```

If you have a vector of character strings, R can count the number of characters in each string:

```{r}
sentence <- c("I", "love", "to", "code", "in", "R.")

nchar(sentence)
```

You can combine vectors of the same data type using the ```c()``` function. This is very useful when you need want to combine vectors. For instance, you may obtain vectors from two different data sets that you want to combine for a single analysis.
```{r}
sentence_2 <- c("But", "Python", "is", "ok", "too.")

c(sentence, sentence_2)
```


Each *data type* has its own unique properties.

Here is a list of the different data types:

* ``int`` stands for integers, or whole numbers.

* ``dbl`` stands for doubles, or numbers with decimals.

* ``chr`` stands for character vectors, or strings.

* ``dttm`` stands for date-times (a date + a time).

* ``lgl`` stands for logical, vectors that contain only TRUE or FALSE.

* ``fctr`` stands for factors, which R uses to represent categorical variables with fixed possible values.

* ``date`` stands for dates.

## typeof()
```typeof()``` is a handy function that returns the data type of your vector. If you're ever in doubt, check your vector with ```typeof()```!

```{r}
dec <- c(3.24, 5.4, 2.04, 6.55)

typeof(dec)
```



## Getting it right

Vectors cannot contain more than one type of data. If you try, R will coerce your vector to a single data type that you weren't intending.

```{r}
mix_vec <- c(1, "two", 3L)

mix_vec

typeof(mix_vec)
```

**Exercise 1**. Answers at the bottom.

You're interested in whether modern Western households stick to the "three meals a day" norm that has persisted since the middle ages. Work demands have increased and you think that people may be sherking a meal in favor of more work time. In a preliminary survey of your apartment complex, you surveyed four households and asked how many meals each member consumes on average each day. 

Complete the following to analyze your preliminary data.

1) Check your data vectors to make sure you are indeed working with numeric data.
```{r}
# hh = household
hh_1 <- c(2, 3, 3)
hh_2 <- c(3, 2, 2, 3)
hh_3 <- c(2, 4)
hh_4 <- c(3, 2)
```


2) Take the average of each household using the ```take_avg()``` function you created earlier. Assign each one to a unique object.
3) Create a new vector of your averages.
4) Visualize your new household averages using the ```barplot()``` function. Just put your object from step 3) inside the function, e.g. ```barplot(hh_averages)```. 
5) Since you're interested in average across households, take the average of your household averages. What does your preliminary survey tell you?


## Exercise Answers

```{r, include=FALSE}
# read in function again
take_avg <- function(x){
  
  # this function takes the same of all values in x. Assumes x is numerical
  s <- sum(x)
  
  # this counts the number of elements in x
  l <- length(x)
  
  # we're dividing the sum of values by the number of values to calculate the mean
  avg <- s / l
  
  # this specifies the value to be returned by the function
  return(avg)
}
```


**Exercise 1**

1)
```{r, collapse=TRUE}
typeof(hh_1)
typeof(hh_2)
typeof(hh_3)
typeof(hh_4)
```

2)
```{r}
hh_avg_1 <- take_avg(hh_1)
hh_avg_2 <- take_avg(hh_2)
hh_avg_3 <- take_avg(hh_3)
hh_avg_4 <- take_avg(hh_4)
```

3) 
```{r}
all_avgs <- c(hh_avg_1, hh_avg_2, hh_avg_3, hh_avg_4)
all_avgs
```

4)
```{r}
barplot(all_avgs)
```

5)
```{r}
take_avg(all_avgs)
```



[\<\<\< Previous](04-vectors.md) | [Next \>\>\>](06-data-structure.md)  
[Glossary](glossary.md)
